---
title: "Asiye Demir Style Guide"
date: "01/30/2021"
author: "Asiye Demir"
output: html_notebook
---

### Description

Rules below show my understanding of R syntax and the process of operation. 
Because I am a beginner in coding, I follow one of the professional's guide 
as an example. Among many other style guides, I like 
[Jean Fan](http://jef.works/R-style-guide/)
guide better. It is plain and provides many basic for a beginner like me. 

### Naming Convention

Descriptive naming is an important part for consistency of the work. 
As some projects cannot be accomplished in one day, after I return to work on R,
I want to be able to remember the names. 

I prefer to use lower case letters instead of uppercase to name objects. 
I do not like underscore (_) in the names, but if I need to be descriptive for 
a combination of variables, I use it. Examples below just for the naming, 
please ignore the mathematical concept.

```{r eval=FALSE, include=FALSE}
reading<- 3 + 5 # single variable or function 
reding_SES<- 3 + 5 #reading scores contingent on SES levels 

```

### Syntax

I prefer to use one space bar indent instead of too close or too wide. Left 
sided arrow stores the vectors or functions to an object. The text or the codes
in the chunks follows the 80 character line that appears on the right side of 
notebook. In the code chunk, the object or variable in the left side of the 
operator does not have a space, but the functions that are right side of the 
arrow has a space.

```{r eval=FALSE, include=FALSE}

death2012<- subset(guns, guns$year == "2012")

More_death2012<- death2012 %>%
  group_by(intent) %>%
  summarize(freq = n()) %>%
  mutate(prop = freq/sum(freq))
```

### Commenting

I usually do not use comments with # in the code chunk unless I am working 
something very complex. If I need to remember what this chunk is for, I put 
some small comments. Other than this, #s for the scrip commenting. 

```{r eval=FALSE, include=FALSE}
#model 5 has college grads + no HS grads variables

model5<- lm(police_ed$Police.Fund ~ police_ed$College.Grads + police_ed$No.HS)
summary(model5)
```









